A meta analysis of tournaments and an evaluation of performance in the Iterated Prisoner's Dilemma.

### 2019-09-10 paper/main.tex
draft of first sections and will re-write evaluation section

The iterated prisoner's dilemma has been used for decades as a powerful model of
behavioural interactions. From the celebrated performance of Tit for Tat, to the
introduction of the zero-determinant family, to the use of sophisticated
strategies such as neural networks, the literature has been exploring the
performance of strategies in the game for years. Most of these strategies
are now accessible due to an open source package. This manuscript make use
of the open source and it's strategy data base to a conducts a meta analysis of
iterated prisoner's dilemma tournaments. The aim is to evaluate the performance
of these strategies and finally answer the question: which is the best strategy
in the game?
\end{abstract}

@@

The iterated prisoner's dilemma is a repeated two player game that models
situations in which self-interest clashes with collective interest. At each turn,
the players simultaneously and independently make a choice between cooperation (C) and
defection (D) whilst their prior interactions matter. The payoffs at each given turn are defined by the matrix,

@@

Since the computer tournaments of R. Axelrod in 1980s several academic papers
are published in the field regarding the performance of strategies in the
iterated prisoner's dilemma. In the 80's following the strong performance of Tit For Tat
in both Axelrod's computer tournaments~\cite{Axelrod1980a, Axelrod1980b}, and
moreover in a series of evolutionary experiments~\cite{Axelrod1981}, the strategy
was thought as the most robust basic strategy in the iterated prisoner's
dilemma. However, the strategy's poor performance in environments with
noise~\cite{Bendor1991, Donninger1986, Molander1985, Hammerstein1984},
made room for other protagonists, such as, Nice and Forgiving~\cite{Bendor1991},
Pavlov~\cite{Nowak1993} and Generous Tit For Tat~\cite{Nowak1992}.

Yet again, in 2012 another set of strategies was introduced as the dominant set
of strategies~\cite{Press2012}. These were called zero-determinant strategies,
and by forcing a linear relationship between the payoffs they can ensure that
they will never receive less than their opponents. However, in~\cite{Harper2017}
a tournament containing over 200 strategies, including the zero-determinant strategies
introduced in~\cite{Press2012}, was performed and none of the zero-determinant
strategies ranked in top spots. Instead, the top ranked strategies were a set of
trained strategies based lookup tables, hidden markov models and finite state
automata.
%TODO reference the archetypes of the archetypes used the IPD?

Thus, the following question is raised here: which are the true dominant
strategies in the iterated prisoner's dilemma?

This manuscript uses the open source package
Axelrod-Python~\cite{axelrodproject} to simulate a large number of computer
tournaments using as many strategies as possible from the literature. The aim is
evaluate the performance of these strategies in a tournament and futhermore,
explore the factors of their success. This is done not for standard round robin
tournaments, but also for noisy, probabilistic ending and noisy probabilistic
ending tournaments.

The structure of this manuscript is as follows. In
Section~\ref{section:data_collection} the data collection is covered. In
Section~\ref{section:top_performances}, the best performed strategies for each
type of tournament and overall are presented.
Section~\ref{section:evaluation_of_performance}, explores the traits which
contribute to good performance and finally in Section~\ref{section:conclusion}
we conclude and summarise the results.

@@

For the purposes of this manuscript a data set containing results on iterate
prisoner's dilemma tournaments has been generated, which is available at. This
was done using the open source package Axelrod-Python~\cite{axelrodproject},
more specifically, version 3.0.0. Axelrod-Python allows for different types of
iterated prisoner's dilemma computer tournaments to be simulated whilst
containing a list of over 190 strategies. Most of these are strategies described
in the literature with a few exceptions being strategies that have been
contributed specifically to the package. Though Axelrod-Python features several
tournament types, this work considers only standard, noisy, probabilistic ending
and noisy probabilistic ending tournaments.

\textbf{Standard tournaments}, are tournaments similar to that of Axelrod's
in~\cite{Axelrod1980a}. There are \(N\) strategies which all play an iterated
game of \(n\) number of turns against each other. Note that self interactions and
a match against a random strategy are not included. Similarly, \textbf{noisy tournaments}
have \(N\) strategies and \(n\) number of turns but at each turn there is a
probability \(p\) that a player's action will be flipped.
\textbf{Probabilistic ending tournaments}, are of size \(N\) and after each turn
a match between strategies ends with a given probability \(e\). Finally,
\textbf{noisy probabilistic ending} tournaments have both a noise probability
\(p\) and an ending probability \(e\). For smoothing the simulated results a
tournament is repeated for \(r\) number of times and the ranks are based on the
average score a strategy achieved and not by number of wins. A summary of
each tournaments' parameters is given in Table~\ref{table:tournaments_parameters}.

\begin{table}[!htbp]
    \begin{center}
        \resizebox{.85\textwidth}{!}{
        \begin{tabular}{lccccc}
    \toprule
    tournament type & number of strategies & turns & noise probability & probability of match ending \\
    \midrule
    standard & $N$  & $n$ & - & - \\
    noisy & $N$  & $n$ & $p$ & - \\
    probabilistic ending & $N$  & - & - & $e$ \\
    noisy probabilistic ending & $N$  & - & $p$ & $e$ \\
    \bottomrule
        \end{tabular}}
    \end{center}
    \caption{Tournament types' parameters.}
    \label{table:tournaments_parameters}
\end{table}

The process of generating data implemented in this manuscript is given by
Algorithm~\ref{algorithm:data_generation}. For each trial a random
size \(N\) is selected, and from the list of 195 strategies
in~\cite{axelrodproject}, a random list of \(N\) strategies is chosen. For the
given list of strategies a standard, noise, probabilistic ending and a noisy
probabilistic ending tournament are performed and repeated \(r\) times.
The parameters for the tournaments as well as the number of repetitions are
selected once for each trial. The parameters and their respective minimum and
maximum values are given by Table~\ref{table:parameters_values}.

@@

\begin{algorithm}[!htbp]
    \setstretch{1.35}
    \ForEach{\text{seed} $\in [0, 12285]$}{
        $N \gets \text{randomly select integer}\in [N_{min}, N_{max}]$\;
        $\text{players} \gets  \text{randomly select $N$ players}$\;
        $r \gets  \text{randomly select integer}\in [r_{min}, r_{max}]$\;
        $n \gets  \text{randomly select integer}\in [n_{min}, n_{max}]$\;
        $p \gets  \text{randomly select float}\in [p_{min}, p_{max}]$\;
        $e \gets   \text{randomly select float}\in [e_{min}, e_{max}]$\;
        \vspace{0.4cm}
        $\text{result standard}$ $\gets$ Axelrod.tournament$(\text{players}, n, r)$\;
        $\text{result noisy}$ $\gets$ Axelrod.tournament$(\text{players}, n, p, r)$\;
        $\text{result probabilistic ending}$ $\gets$ Axelrod.tournament$(\text{players}, e, r)$\;
        $\text{result noisy probabilistic ending}$ $\gets$ Axelrod.tournament$(\text{players}, p, e, r)$\;

@@

\begin{table}[!htbp]
    \begin{center}
        \resizebox{.6\textwidth}{!}{
        \begin{tabular}{lcccc}
    \toprule
    parameter & parameter explanation &   min value & max value \\
    \midrule
    $N$ & number of strategies  & 3 & 195 \\
    $r$ & number of repetitions  & 10 & 100 \\
    $n$ & number of turns      & 1 & 200 \\
    $p$ & probability of flipping action at each turn  & 0 & 1   \\
    $e$ & probability of match ending in the next turn & 0 & 1   \\
    \bottomrule
        \end{tabular}}
    \end{center}
    \caption{Data generation parameters' values}
    \label{table:parameters_values}
\end{table}

A total of 12,285 trials of Algorithm~\ref{algorithm:data_generation} have been
performed. For each trial the results for 4 different tournaments were collected,
thus a total of 49,140 $(12,285 \times 4)$ tournament results have been
retrieved. Each tournament outputs a result summary in the form of
Table~\ref{table:output_result}. The result summary has a length \(N\) because
each row contains information for each strategy that participated in the tournament.
The information include the strategy's rank $(R)$, median score, the rate with
which the strategy cooperated $(C_r)$, it's wins and the probability that the strategy
cooperated in the opening move. Moreover the summary result includes the rates
of a strategy being in any of the four states ($CC, CD, DC, DD$), and the rate
of which the strategy cooperated after each state.

\newcolumntype{g}{>{\columncolor{Gray}}c}
\begin{table}[!htbp]
    \begin{center}
    \resizebox{\textwidth}{!}{
    \begin{tabular}{ccccccgcgcgcgcg}
    \toprule 
    & & & & & &   \multicolumn{8}{g}{Rates}  \\
    Rank $(R)$ & Name & Median score & Cooperation rating $(C_r)$ & Win & Initial C &
    CC & CD & DC & DD & CC to C & CD to C & DC to C & DD to C \\
    0 &  EvolvedLookerUp2 2 2 & 2.97 & 0.705 & 28.0 & 1.0 & 0.639 & 0.066 & 0.189 &
    0.106 & 0.836 & 0.481 & 0.568 & 0.8 \\
    1 &  Evolved FSM 16 Noise 05 & 2.875 & 0.697 & 21.0 & 1.0 & 0.676 & 
    0.020 & 0.135 & 0.168 & 0.985 & 0.571 & 0.392 & 0.07 \\
    2 & PSO Gambler 1 1 1 & 2.874 & 0.684 &  23.0 &     1.0 &    0.651 &    0.034 &    0.152 &    0.164
    & 1.000 & 0.283 & 0.000 & 0.136 \\
    3 &  PSO Gambler Mem1 &  2.861 &        0.706 &  23.0 &      1.0 &    0.663
    &    0.042 &    0.145 &    0.150 &  1.000 &  0.510 &  0.000 &  0.122 \\
    4 &          Winner12 &  2.835 &        0.682 &  20.0 &      1.0 &
    0.651 &    0.031 &    0.141 &    0.177 &  1.000 &  0.441 &  0.000 &  0.462 \\
    $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ &
    $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ \\
    \bottomrule
    \end{tabular}}
\end{center}
\caption{Output result.}\label{table:output_result}
\end{table}

Several other measures have been manually calculated/retrieved and have been
included in the data set. These include properties of the strategies as well as
measures regarding the tournament, such as the parameters of
Table~\ref{table:parameters_values}. All these measures are summarized and
described Table~\ref{table:manual_measures}.

\newcolumntype{g}{>{\columncolor{Gray}}c}
\begin{table}[h]
    \begin{center}
    \resizebox{\textwidth}{!}{
    \begin{tabular}{gcgcgc}
    \toprule
    measure & measure explanation &  source & value type & min value & max value \\
    \midrule
normalized rank $(r)$ & The rank of the strategy divided by the tournament size & result summary & float & 0 & 1 \\
memory depth &  Strategy's memory size & strategy classifier from~\cite{axelrodproject} & integer & 0 & $\infty$ \\
stochastic  &  If a strategy is stochastic & strategy classifier from~\cite{axelrodproject} & boolean  & False &  True \\
makes use of game &  If a strategy makes used of the game information & strategy classifier from~\cite{axelrodproject} & boolean  & False &  True \\
makes use of length &  If a strategy makes used of the number of turns & strategy classifier from~\cite{axelrodproject} & boolean  & False &  True \\
SSE & A measure of how far a strategy is from extortionate behaviour & method described in [ref] & float & 0 & 1 \\
max cooperating rate $(C_{\text{max}})$  & The biggest cooperating rate in the tournament & result summary & float & 0 & 1 \\
min cooperating rate $(C_{\text{min}})$ & The smallest cooperating rate in the tournament & result summary & float & 0 & 1 \\
median cooperating rate $(C_{\text{median}})$ & The median cooperating rate in the tournament & result summary & float & 0 & 1 \\
mean cooperating rate $(C_{\text{mean}})$ & The mean cooperating rate in the tournament & result summary & float & 0 & 1 \\
$C_r$ / $C_{\text{max}}$ & A strategy's cooperating rate divided by the maximum & manually & float & 0 & 1 \\
$C_r$ / $C_{\text{min}}$ & A strategy's cooperating rate divided by the minimum & manually & float & 0 & 1 \\
$C_r$ / $C_{\text{median}}$ & A strategy's cooperating rate divided by the median & manually & float & 0 & 1 \\
$C_r$ / $C_{\text{mean}}$ & A strategy's cooperating rate divided by the mean & manually & float & 0 & 1 \\
    \bottomrule
        \end{tabular}}
    \end{center}
    \caption{Manually calculated/retrieved measures.}
    \label{table:manual_measures}
    \end{table}

These measures will be used in more detail in the following sections.

% Might move these measures to the sections in which they are actual used as
% it feels like an overload of info in this section.

The data collection and the output from the generated tournaments have been
covered in this section. The measures that will be used in the analysis of the
following sections have been presented. These are also summarised in Table in
the Appendix. In the following Section, the top performances at each tournament
are overall are presented.

@@

This section evaluates the performance of 186 strategies. These can be find in
the Appendix. The performance of each strategy will be evaluated for each type
of tournament independently. In Section~\ref{subsection:standard_tournament} the
strategies are evaluated on their performance in standard tournaments, in
Section~\ref{subsection:noisy_tournament} in noisy, and the rest in
Sections~\ref{subsection:probend_tournament}
and~\ref{subsection:noisy_probend_tournament}.

Each tournament could have participated in multiple tournaments of the same
type. On average a strategy participated in 5690 different tournaments of each
type. For example Tit For Tat has participated in a total of 5569 standard
tournaments. The strategy's normalised ranks distribution is given in
Figure~\ref{fig:tit_for_tat_r_distribution}. As a result, of the multiple entries of
strategies their performance in this manuscript is evaluated based on the median
normalised rank denoted as \(\bar{r}\). Note that a value of \(\bar{r} = 0\)
means that a strategy the winner of the tournament where a value of \(\bar{r} = 1\)
means that the strategy came last.

\begin{figure}
    \centering
    \includegraphics[width=.5\textwidth]{../images/tit_for_tat_r_distribution.pdf}
    \caption{Tit For Tat's $r$ distribution for standard tournaments.}
    \label{fig:tit_for_tat_r_distribution}
\end{figure}

Following the analysis of each tournament type, in Section~\ref{subsrction:overall}
the results are merged and the strategies are evaluated in their overall performances
over all the different tournaments.

\subsection{Standard tournaments}\label{subsection:standard_tournament}

The results presented in this section are based on $12,285$ different standard
tournaments. Each tournament has on average 122 participants, a length of
100 turns and was repeated 55 times.

The top 15 performances in standard tournaments are given by
Table~\ref{table:std_results}. The 10 out of the 15 are strategies introduced
in~\cite{Harper2017}. These have been trained using FSM (finite state automata),
HMM (hidden markov models), ANN (artificial neural networks), lookup tables
(LookerUp) and stochastic lookup tables (Gambler). Specifically, the have been
trained using the strategy list of~\cite{axelrodproject} in standard
tournaments. Thus their performance was to be expected. DoubleCrosser, and Fool
Me Once, are not from the literature but from~\cite{axelrodproject}.
DoubleCrosser is a strategy that makes use of the length of the match because is
set to defect on the last two rounds. The strategy is expected to not perform as
well in probabilistic ending tournaments. Finally, Winner 12~\cite{Mathieu2017}
and DBS are both from the the literature. DBS~\cite{Au2006} is strategy
specifically designed for noisy environments, however, it ranks highly only in
standard ones.

\begin{table}[!htbp]
    \centering
    \resizebox{.3\textwidth}{!}{
    \input{performance_standard_table.tex}}
    \caption{Standard top performances}\label{table:std_results}
\end{table}

The distributions of $\bar{r}$ for top ranked strategies are given in
Figure~\ref{fig:std_results}. The distributions are skewed towards zero and the
highest median is at 0.075. These strategies would mostly rank high when they
participated in a tournament. This is not true for the rest of the tournament
types. In the following sections $\bar{r}$ and subsequently the performances are
more variant.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.8\textwidth]{../images/performance_standard.pdf}
    \caption{$\bar{r}$ distributions of top 15 strategies.}\label{fig:std_results}
\end{figure}

\subsection{Noisy tournaments}\label{subsection:noisy_tournament}

Similarly to Section~\ref{subsection:standard_tournament} the results presented
here are based on $12,285$ different noisy tournaments. The strategies have been
ranked based on their \(\bar{r}\) and the top 15 are given by
Table~\ref{table:noisy_results}. The distributions of their corresponding
\(\bar{r}\) is also given by Figure~\ref{fig:noisy_results}.

The top strategies include two well known strategies Tit For 2
Tats~\cite{Axelrod1980b} and Hard Tit For 2 Tats~\cite{Stewart2012} which are
strategies that will defect only after the have received two defections from the
opponent. The Retaliate strategies are set of strategies from~\cite{axelrodproject}
that start by cooperating but will retaliate once the opponent's wins and
defections surpass a curtain threshold. Strategies ShortMem~\cite{Carvalho2013},
Grumpy, $e$ and $\phi$ are strategies that make decisions based on the
cooperations to defections. Cycler Hunter tries to extort strategies that play
cyclically and Risky QLearn uses a Q learning algorithm. Notably a very simple
strategy, Cooperator, has been ranked third.

\begin{table}[!htbp]
    \centering
    \resizebox{.28\textwidth}{!}{
    \input{performance_noise_table.tex}}
    \caption{Noisy top performances}\label{table:noisy_results}
\end{table}

From Figure~\ref{fig:std_results} it is evident that the normalised rank distributions
in noisy environments are more variant and have higher median values compared
to standard tournament. Moreover, the distributions are skewed both towards
0 and 1 which indicate that the top ranked strategies ranked first and last
in several tournaments. However, the low values of the median shows that the
have won most of the tournaments.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.8\textwidth]{../images/performance_noise.pdf}
    \caption{\(\bar{r}\) distributions for best performed strategies in noisy tournaments.}
    \label{fig:noisy_results}
\end{figure}

\subsection{Probabilistic ending tournaments}\label{subsection:probend_tournament}

In this section the performances in probabilistic ending tournaments are discussed.
The 15 top ranked strategies are given by Table~\ref{table:prob_end_results}
and their respective $\bar{r}$ distributions in Figure~\ref{fig:probend_results}.

Fortress 3, Fortress 4 (both introduced in~\cite{Ashlock2006}),
Raider~\cite{Ashlock2014} and Solution B1~\cite{Ashlock2014} are strategies
based on finite state automata introduced by Daniel and Wendy Ashlock. These
strategies have been evolved using reinformed learning. However, there were
trained to maximise their payoffs in tournaments with fixed turns, 150
specifically, and not in probabilistic ending tournaments.

>>>>

The Iterated Prisoner's Dilemma has been used for decades as a powerful model of
behavioural interactions. From the celebrated performance of Tit for Tat, to the
introduction of the zero-determinant strategies, to the use of sophisticated
structures such as neural networks, the literature has been exploring the
performance of strategies in the game for years. Most of these strategies are
now accessible due to an open source package; Axelrod-Python. This manuscript
make use of Axelrod-Python to conduct a meta analysis of 40000 Iterated
Prisoner's Dilemma tournaments. The aim is to evaluate the performance of
numerous strategies and finally answer the explore the factors of success in the
game.
\end{abstract}

@@

The Iterated Prisoner's Dilemma (IPD) is a repeated two player game that models
situations in which self-interest clashes with collective interest. At each turn,
the players simultaneously and independently make a choice between cooperation (C) and
defection (D) whilst having memory of the prior interactions.
The payoffs at each given turn are defined by the matrix,

@@

Since the computer tournaments of R. Axelrod in 1980s several academic papers
are published in the field regarding the performance of strategies in the
IPD. In the 80's following the strong performance of Tit For Tat
in both Axelrod's computer tournaments~\cite{Axelrod1980a, Axelrod1980b}, and
moreover in a series of evolutionary experiments~\cite{Axelrod1981}, the strategy
was thought as the most robust basic strategy in the iterated prisoner's
dilemma. However, the strategy's poor performance in environments with
noise~\cite{Bendor1991, Donninger1986, Molander1985, Hammerstein1984},
made room for other protagonists, such as, Nice and Forgiving~\cite{Bendor1991},
Pavlov~\cite{Nowak1993} and Generous Tit For Tat~\cite{Nowak1992}.

In 2004, the Anniversary Iterated Prisoner Dilemma Tournament was run with 233
entries. The winning strategies were based on a mechanism of teams. A team from
Southampton University took advantage of the fact that a participant was allowed
to submit multiple strategies. They submitted a total of 60 of strategies that
could recognised each other and colluded to increase one members
score~\cite{J.P.Delahaye1993Lp, J.P.Delahaye1995LIeP, A.Rogers2007Ctpw}.

Yet again, in 2012 another set of strategies was introduced as the dominant set
of strategies~\cite{Press2012}. These were called zero-determinant strategies,
and by forcing a linear relationship between the payoffs they can ensure that
they will never receive less than their opponents. However, in~\cite{Harper2017}
a tournament containing over 200 strategies, zero-determinant, was performed and
none of the zero-determinant strategies ranked in top spots. Instead, the top
ranked strategies were a set of evolved strategies based lookup tables, hidden
markov models and finite state automata.
%TODO reference the archetypes of the archetypes used the IPD?

Thus, the following question is raised here: which are the true dominant
strategies in the iterated prisoner's dilemma?
This manuscript uses the open source package
Axelrod-Python~\cite{axelrodproject} to simulate a large number of computer
tournaments using as many strategies as possible from the literature. The aim is
to evaluate the performance of these strategies in a tournament and futhermore,
explore the factors of their success. This is done not for standard round robin
tournaments, but also for noisy, probabilistic ending and noisy probabilistic
ending tournaments.

The different tournaments and the data generating process are covered in
Section~\ref{section:data_collection}. Section~\ref{section:top_performances},
covers the best performed strategies for each type of tournament and overall.
Section~\ref{section:evaluation_of_performance}, explores the traits which
contribute to good performance and finally in Section~\ref{section:conclusion}
the results are discussed and summarised.

@@

For the purposes of this manuscript a data set containing results on IPD tournaments
has been generated and is available at. This
was done using the open source package Axelrod-Python~\cite{axelrodproject},
more specifically, version 3.0.0. Axelrod-Python allows for different types of
IPD computer tournaments to be simulated whilst
containing a list of over 180 strategies. Most of these are strategies described
in the literature with a few exceptions being strategies that have been
contributed specifically to the package. Though Axelrod-Python features several
tournament types, this work considers only standard, noisy, probabilistic ending
and noisy probabilistic ending tournaments.

\textbf{Standard tournaments}, are tournaments similar to that of Axelrod's
in~\cite{Axelrod1980a}. There are \(N\) strategies which all play an iterated
game of \(n\) number of turns against each other. Note that self interactions
and a match against a random strategy are not included. Similarly, \textbf{noisy
tournaments} have \(N\) strategies and \(n\) number of turns but at each turn
there is a probability \(p\) that a player's action will be flipped.
\textbf{Probabilistic ending tournaments}, are of size \(N\) and after each turn
a match between strategies ends with a given probability \(e\). Finally,
\textbf{noisy probabilistic ending} tournaments have both a noise probability
\(p\) and an ending probability \(e\). For smoothing the simulated results a
tournament is repeated for \(k\) number of times. The winner of each tournament
is based on the average score a strategy achieved and not by number of wins.

%  A summary of
% each tournaments' parameters is given in Table~\ref{table:tournaments_parameters}.

% \begin{table}[!htbp]
%     \begin{center}
%         \resizebox{.85\textwidth}{!}{
%         \begin{tabular}{lcccccc}
%     \toprule
%     tournament type & number of strategies & repetitions & turns & noise probability & probability of match ending \\
%     \midrule
%     standard & $N$ & k & $n$ & - & - \\
%     noisy & $N$  & k & $n$ & $p$ & - \\
%     probabilistic ending & $N$ & k & - & - & $e$ \\
%     noisy probabilistic ending & $N$ & k & - & $p$ & $e$ \\
%     \bottomrule
%         \end{tabular}}
%     \end{center}
%     \caption{Tournament types' parameters.}
%     \label{table:tournaments_parameters}
% \end{table}

The process of generating data implemented in this manuscript is given by
Algorithm~\ref{algorithm:data_generation}. For each trial a random
size \(N\) is selected, and from the list of 186 strategies
in~\cite{axelrodproject}, a random list of \(N\) strategies is chosen. For the
given list of strategies a standard, a noisy, a probabilistic ending and a noisy
probabilistic ending tournament are performed and repeated \(k\) times.
The parameters for the tournaments as well as the number of repetitions are
selected once for each trial. The parameters and their respective minimum and
maximum values are given by Table~\ref{table:parameters_values}.

\begin{table}[!htbp]
    \begin{center}
        \resizebox{.6\textwidth}{!}{
        \begin{tabular}{lcccc}
    \toprule
    parameter & parameter explanation &   min value & max value \\
    \midrule
    $N$ & number of strategies  & 3 & 195 \\
    $k$ & number of repetitions  & 10 & 100 \\
    $n$ & number of turns      & 1 & 200 \\
    $p$ & probability of flipping action at each turn  & 0 & 1   \\
    $e$ & probability of match ending in the next turn & 0 & 1   \\
    \bottomrule
        \end{tabular}}
    \end{center}
    \caption{Data generation parameters' values}
    \label{table:parameters_values}
\end{table}

@@

\begin{algorithm}[!htbp]
    \setstretch{1.35}
    \ForEach{\text{seed} $\in [0, 12285]$}{
        $N \gets \text{randomly select integer}\in [N_{min}, N_{max}]$\;
        $\text{players} \gets  \text{randomly select $N$ players}$\;
        $k \gets  \text{randomly select integer}\in [k_{min}, k_{max}]$\;
        $n \gets  \text{randomly select integer}\in [n_{min}, n_{max}]$\;
        $p \gets  \text{randomly select float}\in [p_{min}, p_{max}]$\;
        $e \gets   \text{randomly select float}\in [e_{min}, e_{max}]$\;
        \vspace{0.4cm}
        $\text{result standard}$ $\gets$ Axelrod.tournament$(\text{players}, n, k)$\;
        $\text{result noisy}$ $\gets$ Axelrod.tournament$(\text{players}, n, p, k)$\;
        $\text{result probabilistic ending}$ $\gets$ Axelrod.tournament$(\text{players}, e, k)$\;
        $\text{result noisy probabilistic ending}$ $\gets$ Axelrod.tournament$(\text{players}, p, e, k)$\;

@@

A total of 12,285 trials of Algorithm~\ref{algorithm:data_generation} have been
performed. For each trial the results for 4 different tournaments were collected,
thus a total of 49,140 $(12,285 \times 4)$ tournament results have been
retrieved. Each tournament outputs a result summary in the form of
Table~\ref{table:output_result}.

The result summary has a length \(N\) because each row contains information for
each strategy that participated in the tournament. The information include the
strategy's rank, median score, the rate with which the strategy cooperated
$(C_r)$, it's wins and the probability that the strategy cooperated in the
opening move. Moreover, the rates of a strategy being in any of the four states
($CC, CD, DC, DD$), and the rate of which the strategy cooperated after each
state.

\newcolumntype{g}{>{\columncolor{Gray}}c}
\begin{table}[!htbp]
    \begin{center}
    \resizebox{\textwidth}{!}{
    \begin{tabular}{ccccccgcgcgcgcg}
    \toprule 
    & & & & & &   \multicolumn{8}{g}{Rates}  \\
    Rank & Name & Median score & Cooperation rating $(C_r)$ & Win & Initial C &
    CC & CD & DC & DD & CC to C & CD to C & DC to C & DD to C \\
    0 &  EvolvedLookerUp2 2 2 & 2.97 & 0.705 & 28.0 & 1.0 & 0.639 & 0.066 & 0.189 &
    0.106 & 0.836 & 0.481 & 0.568 & 0.8 \\
    1 &  Evolved FSM 16 Noise 05 & 2.875 & 0.697 & 21.0 & 1.0 & 0.676 & 
    0.020 & 0.135 & 0.168 & 0.985 & 0.571 & 0.392 & 0.07 \\
    2 & PSO Gambler 1 1 1 & 2.874 & 0.684 &  23.0 &     1.0 &    0.651 &    0.034 &    0.152 &    0.164
    & 1.000 & 0.283 & 0.000 & 0.136 \\
    3 &  PSO Gambler Mem1 &  2.861 &        0.706 &  23.0 &      1.0 &    0.663
    &    0.042 &    0.145 &    0.150 &  1.000 &  0.510 &  0.000 &  0.122 \\
    4 &          Winner12 &  2.835 &        0.682 &  20.0 &      1.0 &
    0.651 &    0.031 &    0.141 &    0.177 &  1.000 &  0.441 &  0.000 &  0.462 \\
    $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ &
    $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ \\
    \bottomrule
    \end{tabular}}
\end{center}
\caption{Output result.}\label{table:output_result}
\end{table}

The normalised rank is a measure which was manually included. The normalised
rank, denoted as $r$, is calculated as a strategy's rank divided by the
tournament's size ($N$). The normalised rank will be used in the next section to
evaluate the performance of strategies.

@@

This section evaluates the performance of 186 strategies which can be found in
the Appendix. The performance of each strategy will be evaluated for each type
of tournament independently. In Section~\ref{subsection:standard_tournament} the
strategies are evaluated on their performance in standard tournaments, in
Section~\ref{subsection:noisy_tournament} in noisy, in
Sections~\ref{subsection:probend_tournament} in probabilistic ending and in
\ref{subsection:noisy_probend_tournament} in noisy probabilistic ending.

Each strategy could have participated in multiple tournaments of the same type
(on average each participated in 5690 different tournaments). For example Tit
For Tat has participated in a total of 5569 standard tournaments. The strategy's
normalised rank distribution in different tournament is given in
Figure~\ref{fig:tit_for_tat_r_distribution}. As a result, of the multiple
entries of strategies their performance is evaluated based on the median
normalised rank denoted as \(\bar{r}\). A value of \(\bar{r} = 0\) corresponds
to a strategy winning the tournament where a a value of \(\bar{r} = 1\)
corresponds to the strategy coming last.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.8\textwidth]{../images/tit_for_tat_r_distributions.pdf}
    \caption{Tit For Tat's $r$ distribution in tournaments.}
    \label{fig:tit_for_tat_r_distribution}
\end{figure}

Following the analysis of each tournament type, in Section~\ref{subsrction:overall}
the results are merged and the strategies are evaluated in their overall performances
over all simulated tournaments of this work.

\subsection{Top ranked strategies in: Standard tournaments}\label{subsection:standard_tournament}

The results presented in this section are based on $12,285$ different standard
tournaments.
The top 15 performances in standard tournaments are given in
Table~\ref{table:std_results}. Ten of these are strategies introduced
in~\cite{Harper2017}. These have been trained using reinforcement learning
algorithms and are based on finite state automata (FSM), hidden markov models
(HMM), artificial neural networks (ANN), lookup tables (LookerUp) and stochastic
lookup tables (Gambler). Specifically, the have been trained against the
strategy list of~\cite{axelrodproject} in standard tournaments. Thus their
performance is to be expected. DoubleCrosser, and Fool Me Once, are not from
the literature but from~\cite{axelrodproject}. DoubleCrosser is a strategy that
makes use of the length of the match because is set to defect on the last two
rounds. The strategy is expected to not perform as well in probabilistic ending
tournaments. Finally, Winner 12~\cite{Mathieu2017} and DBS are both from the the
literature. DBS~\cite{Au2006} is strategy specifically designed for noisy
environments, however, it ranks highly only in standard ones.

\begin{table}[!htbp]
    \centering
    \resizebox{.28\textwidth}{!}{
    \input{performance_standard_table.tex}}
    \caption{Standard top performances}\label{table:std_results}
\end{table}

Figure~\ref{fig:std_results} gives the distributions of $\bar{r}$ for the top
ranked strategies. The distributions are skewed towards zero and the highest
median is at 0.075. This indicates that the top ranked strategies are dominating
strategies in standard tournaments. They are very likely to perform well in
standard tournament despite the number of opponents, the opponents, the turns
etc. This does not hold for all the tournament types as it will be discussed
in the following sections. The next section present the results of noisy
tournaments.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.7\textwidth]{../images/performance_standard.pdf}
    \caption{$\bar{r}$ distributions of top 15 strategies.}\label{fig:std_results}
\end{figure}

\subsection{Top ranked strategies in: Noisy tournaments}\label{subsection:noisy_tournament}

Similarly to Section~\ref{subsection:standard_tournament} the results presented
here are based on $12,285$ different noisy tournaments. The strategies have been
ranked based on their \(\bar{r}\) and the top 15 are given by
Table~\ref{table:noisy_results}. The distributions of their corresponding
\(\bar{r}\) is also given in Figure~\ref{fig:noisy_results}.

The top strategies include the two strategies, Tit For 2
Tats~\cite{Axelrod1980b} and Hard Tit For 2 Tats~\cite{Stewart2012}, which are
strategies that will defect only after the have received two defections from the
opponent. The Retaliate strategies are set of strategies
from~\cite{axelrodproject} that start by cooperating but will retaliate once the
opponent's wins and defections surpass a curtain threshold.
ShortMem~\cite{Carvalho2013}, Grumpy, $e$ and $\phi$ are strategies that make
decisions based on the cooperations to defections ratio. In $5^{\text{th}}$ and
$6^{\text{th}}$ place are the strategies Cycler Hunter and  Risky QLearn. Cycler
Hunter tries to extort strategies that play cyclically and Risky QLearn uses a Q
learning algorithm. Notably, a deterministic and one of the most simple
strategies in game is ranked $3^{\text{rd}}$. That is Cooperator, a strategy
that just cooperates.

\begin{table}[!htbp]
    \centering
    \resizebox{.25\textwidth}{!}{
    \input{performance_noise_table.tex}}
    \caption{Noisy top performances}\label{table:noisy_results}
\end{table}

From Figure~\ref{fig:std_results} it is evident that the normalised rank distributions
in noisy environments are more variant and have higher median values compared
to standard tournaments. The distributions are skewed both towards
0 and 1 which indicates that though the top ranked strategies mainly
performed well (medians $< 0.3$) there are several tournaments that they performed
worse than the 60\% of the participants.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.7\textwidth]{../images/performance_noise.pdf}
    \caption{\(\bar{r}\) distributions for best performed strategies in noisy tournaments.}
    \label{fig:noisy_results}
\end{figure}

\subsection{Top ranked strategies: Probabilistic ending tournaments}\label{subsection:probend_tournament}

In this section the performances in probabilistic ending tournaments are
discussed. The 15 top ranked strategies are given by
Table~\ref{table:prob_end_results} and Figure~\ref{fig:probend_results} gives
their respective $\bar{r}$ distributions.
Fortress 3, Fortress 4 (both introduced in~\cite{Ashlock2006}),
Raider~\cite{Ashlock2014} and Solution B1~\cite{Ashlock2014} are strategies
based on finite state automata introduced by Daniel and Wendy Ashlock. These
strategies have been evolved using reinformed learning, however, there were
trained to maximise their payoffs in tournaments with fixed turns (150
specifically) and not in probabilistic ending ones.

In probabilistic ending tournaments it appears that the top ranks are mostly
occupied by defecting strategies which include Better and Better, Gradual
Killer, Hard Prober (all from ~\cite{prison}), Bully (Reverse Tit For
Tat)~\cite{Nachbar1992} and Defector. Thus, it's surprisingly that EasyGo and
Fool Me Forever are ranked $14^{\text{th}}$ and $15^{\text{th}}$. These
strategies are actually the same; they will defect until their opponent defect,
then they will cooperate until the end. Both strategies have repeatedly ranked highly
as shown in Figure~\ref{fig:probend_results} and there are cases for which they
were the winners of tournaments.

\begin{table}[!htbp]
    \centering
    \resizebox{.25\textwidth}{!}{
    \input{performance_probend_table.tex}}
    \caption{Probabilistic ending top performances}\label{table:prob_end_results}
\end{table}
